---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bis620.2022

<!-- badges: start -->
[![R-CMD-check](https://github.com/FrankWu520/bis620.2022/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/FrankWu520/bis620.2022/actions/workflows/R-CMD-check.yaml)
[![test-coverage](https://github.com/FrankWu520/bis620.2022/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/FrankWu520/bis620.2022/actions/workflows/test-coverage.yaml)
[![lint](https://github.com/FrankWu520/bis620.2022/actions/workflows/lint.yaml/badge.svg)](https://github.com/FrankWu520/bis620.2022/actions/workflows/lint.yaml)
<!-- badges: end -->

The goal of bis620.2022 is to get the spectral signature of accelerometry data and plot UKBiobank accelerometry data. I created this package to capture functions, data, and documentation for BIS620 class in Fall 2022.

Using the function `spectral_signature()` in `spectral.R`, you can obtain the spectral signature of UKBiobank data, which is calculated by taking the modulus of the Fourier coefficients of the signal. The function takes a data setwith columns `X`, `Y`, `Z`, and `time`, and returns a data frame with the modulus of the Fourier coefficients for the X, Y, and Z channels.

With the function `accel_plot()` in `plot.R`, you can plot UKBiobank accelerometry data. The function takes a data set that is assumed to have a `time`, `X`, `Y`, and `Z` column along with either a `time` or `freq` column, and returns a a time-series plot faceted by axis.


## Installation

You can install the development version of bis620.2022 from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("FrankWu520/bis620.2022")
```

## Obtain Spectral Signature

This example shows you how to get the spectral signature of the accelerometry data:

```{r}
library(bis620.2022)
data(ukb_accel)
spec_sig <- spectral_signature(ukb_accel[1:100, ], take_log = TRUE)
spec_sig
```

## Plot Accelerometry Data

This example shows you how to create an accelerometry plot for the accelerometry data:

```{r}
library(bis620.2022)
data(ukb_accel)
accel_plot(ukb_accel[1:1000, ])
```


This example shows you how to create an accelerometry plot for the spectral signature of the accelerometry data:
```{r}
library(bis620.2022)
data(ukb_accel)
spec_sig <- spectral_signature(ukb_accel[1:100, ], take_log = TRUE)
accel_plot(spec_sig)
```
